Exploring Constraint Grammar as a SAT problem

### 2015-03-26 paper/nodalida2015.tex
change structure

\title{Towards something :-D}

@@

\begin{document}
\maketitle
\begin{abstract}
We implement Constraint Grammar as a Boolean satisfiability problem.

We experiment with different strategies to maximise the number of instances to apply rules;
this will also lead to different orderings of the rules and ways of solving conflicts.

\end{abstract}

\section{Introduction}
Constraint Grammar (CG) was first introduced by \cite{karlsson1995constraint}. 
It disambiguates output by morphological analyser, 
and has been used for many tasks in computational linguistics, such as POS tagging,
surface syntax and machine translation.
The current state-of-the-art parser VISL CG-3 is implemented (...);
in addition there are a number of finite-state based implementations \cite{nemeskey14}.

We represent CG as a Boolean satisfiability problem using the 
open-source SAT solver MiniSat \cite{een04sat}.
Our implementation is inspired by approaches like \cite{lager98,lager_nivre01},
but we try to emulate the CG formalism more closely than the previous logic-based works.

% NP-complete -- overkill for CG? But gives other benefits.

Works from the late 90s and early 00s mention that the method is very slow;
however, advances in SAT solving technique

\section{CG as a SAT problem}
\label{sect:pdf}

SAT solving is based on a technique called unit propagation:
a set of complex Boolean clauses is simplified, starting from unit
clauses, which consist of just a single variable, and working up to a
solution, where all variables in the set have a True or False value.
We model an ambiguous sentence as a set of clauses, where each analysis is a variable, 
and constraint rules form clauses operating on those variables.

\begin{itemize}
\item The condition of being unambiguously tagged becomes less necessary.
We add many clauses, knowing that some of them end up being false.
\item Rules are unordered. 
\item Rules interact with each other and disambiguate more than same rules in the standard implementation.
\end{itemize}

% Let us demonstrate the approach with the example sentence 
% \emph{the bear sleeps}. After applying morphological tagger, we get the following analyses:

% \begin{verbatim}
% "<the>"
%         "the" det
% "<bear>"
%         "bear" n sg
%         "bear" v pl
% "<sleeps>"
%         "sleep" n pl
%         "sleep" v sg p3
% \end{verbatim}

%We make each analysis a variable; let us call them \texttt{d1}, \texttt{n2}, \texttt{v2}, \texttt{n3} and \texttt{v3} by the proposed POS tag and the position in the sentence.
%For instance, \texttt{n2} is a hypothesis that the correct analysis 
%for the word form \emph{bear}' is \texttt{"bear" n pl}.
%Before applying rules, we must make sure that in each position
%at least one of the analyses is true. We add our first disjunctions:

%\begin{itemize}
%\item [] $d1$ \\ $n2 \vee v2$ \\ $n3 \vee v3$
%\end{itemize}

%Let us add two rules:
%\begin{itemize}
%\item [] \texttt{REMOVE v IF -1 det} \\ 
%            \texttt{REMOVE n IF -1 n} 
%\end{itemize}

%The rules are applied as implications: \emph{if ``the'' is a determiner, then ``bear'' cannot be a verb}.

%\begin{itemize}
%\item [] $d1 \Rightarrow \neg v2$ \\
%           $n2   \Rightarrow \neg n3$
%\end{itemize}

\subsection{Rule ordering}
\label{ssec:ordering}

@@

\subsection{Rule interaction}

The following example illustrates the interaction of rules to produce the desired outcome. We have an ambiguous passage in Spanish, with the following analyses:

\begin{verbatim}
"<la>"
        "el" det def f sg
        "lo" prn pro p3 f sg
"<casa>"
        "casa" n f sg
        "casar" vblex pri p3 sg
        "casar" vblex imp p2 sg
\end{verbatim}

We add the following rules to disambiguate the word \emph{la}:
\begin{itemize}
\item [] \texttt{REMOVE prn IF (1 n) ;} \\
             \texttt{REMOVE det IF (1 vblex) ;}
\end{itemize}

Standard CG will apply first the rule that removes the pronoun reading, because it comes first in the rule file.
The second rule will not fire, because it would remove the last reading for the word \emph{la}, and that is not allowed. To be more cautious, we could require the word at position +1 be disambiguated fully (\texttt{1C} instead of \texttt{1}), but in that case, neither of them would fire. The ambiguity between the readings of \emph{casa} is left untouched by these rules.

With the SAT clauses, we get two solutions: ``la\texttt{:det} casa\texttt{:n}''  and ``la\texttt{:prn} casa\texttt{:v}''. The interaction of the rules regarding \emph{la}  disambiguates \emph{casa} for free, and ordering of the rules doesn't matter. Since we don't have other rules nor further context, we still get two solutions, but there are no illegal combinations (\texttt{det v} or \texttt{prn n}) in one solution.

@@

We took small constraint grammars (\textless{}500 rules) for Spanish and English, and disambiguated a number of texts from Project Gutenberg. Table~\ref{table:results} shows how much the results agree with the result given by VISL CG3 parser with the same grammar and text.

@@

The tendency of needing less rules to disambiguate shows also in a larger context with more complex rules.
In many of the differences, SATCG disambiguated more. (TODO counts)

>>>>

\title{Exploring Constraint Grammar as a SAT problem}

@@

\begin{document}
\maketitle

%\begin{abstract}
%We implement Constraint Grammar as a Boolean satisfiability problem.
%We experiment with different strategies to maximise the number of instances to apply rules;
%this will also lead to different orderings of the rules and ways of solving conflicts.
%\end{abstract}

\section{Introduction}
%Constraint Grammar (CG) was first introduced by \cite{karlsson1995constraint}. 
%It disambiguates output by morphological analyser, 
%and has been used for many tasks in computational linguistics, such as POS tagging,
%surface syntax and machine translation.
%The current state-of-the-art parser VISL CG-3 is implemented (...);
%in addition there are a number of finite-state based implementations \cite{nemeskey14}.

We represent Constraint Grammar \cite{karlsson1995constraint} 
as a Boolean satisfiability (SAT) problem.
%This increases the computational complexity, 
%but we are exploring ways in which it could simplify the grammar writing:
%our experiments show that less rules are needed, and features like ordering of the rules
%and requirement of being unambiguously tagged become irrelevant.
% Our implementation is inspired by \cite{lager98,lager_nivre01},
%but we try to emulate the CG formalism more closely than the previous logic-based works.
Let us demonstrate our approach with the following example in Spanish.

\begin{verbatim}
"<la>"
        "el" det def f sg
        "lo" prn pro p3 f sg
"<casa>"
        "casa" n f sg
        "casar" vblex pri p3 sg
\end{verbatim}

The ambiguous passage can be either a noun phrase, \emph{la\texttt{:det} casa\texttt{:n}} 
`the house'  or a verb phrase \emph{la\texttt{:prn} casa\texttt{:v}} `(he/she) marries her'. 
We add the following rules:

\begin{itemize}
\item [] \texttt{REMOVE prn IF (1 n) ;} \\
             \texttt{REMOVE det IF (1 vblex) ;}
\end{itemize}

Standard CG will apply one\footnote{To be more cautious, we could require the word at position +1 be disambiguated fully (\texttt{1C} instead of \texttt{1}), but in that case, 
neither of the rules would be applied.} of the rules to the word \emph{la}; 
either the one that comes first, or by some other heuristic. 
The second rule will not fire, because it would remove the last reading. 
Both readings of \emph{casa} are left untouched by these rules.

SAT solver 
performs a search, 
and starts building possible models that satisfy both constraints. 
In addition to the given constraints, we have default rules to emulate
the CG principles: an analysis is true if no rule affects it,
and at least one analysis for each word is true---the notion of ``last'' is not applicable.
With these constraints, we get two solutions: \texttt{det n} and \texttt{prn v}. 
The interaction of the rules regarding \emph{la}  disambiguates \emph{casa} 
for free, and ordering of the rules doesn't matter. 
%Without more context or additional rules we cannot fully disambiguate the passage,
%but unlike with standard CG, we get only legal combinations in one solution.

%If we add a word, \emph{la casa grande} `the big house' 
%and a rule that removes verb if followed by an adjective
%\texttt{REMOVE (vblex) IF (1 (adj)) ;}
%we will get a unique solution with \texttt{det n adj}.
%Standard CG will also remove the verb reading, 
%but the choice of the word \emph{la} doesn't contribute to

The most important differences between the traditional and the SAT-based approach are summarised below:

\begin{itemize}
\item The condition of being unambiguously tagged becomes irrelevant.
Rather than waiting for a word to get disambiguated, the SAT solver starts by 
making assumptions (e.g. ``\emph{casa} is a noun'') and working under them,
discarding the assumption if it doesn't lead to a model that satisfies all constraints.

\item Rules are unordered. 
\item Rules interact with each other and disambiguate more than same rules in the standard implementation.
\end{itemize}

%\section{CG as a SAT problem}
%\label{sect:satcg}

%SAT solving is based on a technique called unit propagation:
%a set of complex Boolean clauses is simplified, starting from unit
%clauses, which consist of just a single variable, and working up to a
%solution, where all variables in the set have a True or False value.
%We model an ambiguous sentence as a set of clauses, where each analysis is a variable, 
%and constraint rules form clauses operating on those variables.

\section{Rule ordering}
\label{sec:ordering}

@@

@@

\paragraph{Time}

SAT is NP-complete, whereas the current implementations of CG are polynomial, 
but with advances in SAT solving techniques, the performance isn't as catastrophically bad as in 90s-00s.
We used the open-source SAT solver MiniSat \cite{een04sat}.
We do foreach sentence: foreach rule; 
%whereas VISL CG3 has foreach rule: foreach sentence.
the size of the SAT problem is dependent of the sentence (???).
TODO: time it comparing it to VISL CG3.
Note also that this is a first proof-of-concept implementation and could be optimised.

\paragraph{Performance against VISL CG3}
We took small constraint grammars (\textless{}500 rules) for Spanish and English, and disambiguated a number of texts from Project Gutenberg. Table~\ref{table:results} shows how much the results agree with the result given by VISL CG3 parser with the same grammar and text.

@@

The tendency of needing less rules to disambiguate shows also in a larger context with more complex rules.
SATCG disambiguated more 17 % of the cases for P&P,

\paragraph{Removing rules}

In this experiment, we test the performance of the two implementations by working with a smaller grammar and comparing the result to
